---
title: "Cats vs Dogs in the US"
author: "Katherine Blake, Jorge Diaz, Alejandro Medina and Mark Mueller"
date: "2023-09-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
<BR><b>The data we are analyzing is in regards to pet ownership statistics in the United States. The data includes averages of cats and dogs within state households as well as pet owner's devotion to their pets, whether the owners are willing to stay at a job they dislike for their pets, and whether the owner’s were willing to spend $4k to save their pets.</b><BR>
```{r}
# loading in packages needed for the project
library(here) # used for finding the location of the data set file
library(maps) # used for map visualization code
library(ggplot2) # used for ggplot
library(dplyr) # used to aid in functionality use of visualizations

# loading in the data as pet_data
pet_data = read.csv(here("DATA","pet-ownership-statistics-by-state_Group2.csv"))

```

## Initial Observations

```{r}
head(pet_data) # viewing the first six rows of data
dim(pet_data) # viewing the dimensions of the data
str(pet_data) # viewing the structure of the data
```
<BR><b>As we go through the nature of the data, we can clearly see the dimensions of the data being 49 observations against 9 variables. These variables are the State, Total Household percentage, Dogs Percentage, Average Number of Dogs, Cats percentage, Average Number of cats, Dogs devotion score, Staying at a job they disliked for dog percentage and Would you spend 4k to save a dog percentage. The structure of the data tells us that the State is the only variable classified as a character; while the others are numeric.</b><BR>

## Cleaning the Data
<b>In our data set, Hawaii and Alaska were omitted and Washington DC only has data on the total household percentage. Because of this, we have decided to omit DC completely.</b>
```{r}
# Omitting rows with no data
pet_data <- na.omit(pet_data)
str(pet_data)
```
<b>In pet_data, we now only have the 48 states we want to look at. The next thing we should do is rename the column names to make it possible to generate graphs and easier to work with.</b> 
```{r}
# Changing 'state' to 'region' and shortening column names
colnames(pet_data) <- c("region","householdPerc",	"dogsPerc",	"avgNumDogs",	"catsPerc",	"avgNumCats",	"dogDevScore",	"jobDislikePerc",	"spend4kPerc")
head(pet_data)
```

<b> Finally, we need to make all of the states lower cases so it can be joined with the state map data set. </b>

```{r}
pet_data$region <- tolower(pet_data$region)
```

<b>After cleaning, all of the info on the dataset describe above is still true, other than the names of the columns being changed and dropping the row containing DC. Several column names were changed in order to shorten them, dogDevScore which stands for the dog devotion score, jobDislikePerc stands for the percentage of people that stayed at a job they disliked for their dog, and spend4kPerc stands for the percentage of people that would spend $4,000 to save their dog.</b> 


## Exploratory Data Analysis

### 1.Percentages of households that own any sort of pet

```{r}
pet_data$region <- tolower(pet_data$region)
state_map <- map_data("state")
state_map_pets <- left_join(state_map,pet_data, by = "region")
head(state_map_pets)
household_graph <- ggplot(state_map_pets, aes(x= long, y = lat, group = group)) +
  geom_polygon(aes(fill = householdPerc)) +
  scale_fill_viridis_c() # Create map on total house percentage
household_graph # View graph

```
<BR><B>Most states had a total household percentage of pets higher than 55%. With this, we can know that it is more common for a household in the United States to have a pet than to not. </B><BR>

### 2. Boxplots comparing dog versus cat ownership in households

```{r}
ggplot(pet_data, aes(x=dogsPerc))+ geom_boxplot()
ggplot(pet_data, aes(x=catsPerc))+ geom_boxplot()

```
<BR><b>In both graphs we can see the distribution of higher percentages in the Dogs Percentages data; while the lower percentages are given to the Cats.  The median for dogs would be just below 40; while for cats the median would be just above 25. This median interpretation means that the distribution of dogs' percentages is higher than cats. This tells us the strong preference of people for owning dogs.</b><BR>

### 3. Comparing dog devotion scores versus staying at a job they disliked and spending 4K to save their dog percentages

```{r}
ggplot(pet_data, aes(x = dogDevScore, y = jobDislikePerc))+
  geom_point(color="salmon") +
  stat_smooth(method = "lm", col="lightblue", se=FALSE ,size=1)+ #lm = linear regression
  labs(title = "Percent of Dog Owners who would stay at a job they dislike according to Dog Devotion Score", x = "Dog Devotion Score", y = "jobDislike") +
  theme_dark()

```
<BR><b>There was no clear correlation for the willingness to stay at a disliked job for each state and the dog devotion score. Since there was no correlation, this simply means there was no clear trend to indicate that a higher likelihood of staying at a disliked job because of the dog would be matched by a higher dog devotion score.</b><BR> 

```{r}
ggplot(pet_data, aes(x = dogDevScore, y = spend4kPerc))+
  geom_point(color="salmon") +
  stat_smooth(method = "lm", col="lightblue", se=FALSE ,size=1)+ #lm = linear regression
  labs(title = "Percent of Dog Owners who would spend 4k to save dog compared to Dog Devotion Score", x = "Dog Devotion Score", y = "% Dog Owner Willing to Spend 4k") +
  theme_dark()

```

<BR><b>There was no clear correlation for the dog devotion score or the percent for each state of people willing to spend 4k to save their dog. Since there was no correlation, this simply means there was no clear trend to indicate that a higher likelihood of spending 4k on their dog would be matched by a higher dog devotion score.</b><BR> 

### 4. Comparing spacial correlation with areas where pet owners were willing to spend $4k to save their dog. 

<b>For this section we pulled from another dataset to get the median household income per state. This was found here: https://worldpopulationreview.com/state-rankings/median-household-income-by-state</b>

```{r}
# Loading in the income into med_income
med_income <- read.csv(here("DATA","median-household-income-by-state-[updated-august-2023].csv"))
str(med_income)
```

<b>Once again this data must be cleaned because it has data for DC, Alaska and Hawaii, and the columns need to be renamed to allow for it to be made into a map.</b>

```{r}
#Removes Alaska, Hawaii, and DC
med_income <- subset(med_income, state != "Alaska")
med_income <- subset(med_income, state != "Hawaii")
med_income <- subset(med_income, state != "District of Columbia")

# Renaming the columns for easier usage
colnames(med_income) <- c("region","income")

# Making all state names lower case
med_income$region <- tolower(med_income$region)
str(med_income)
```

<b>Now that the data has been cleaned, we can use it to create a map and compare it to. First, here is the map seeing the percentage of people that would spend 4K to save their dog.</b>

```{r}
# Joining data set with state map
state_map <- map_data("state")
state_map_pets <- left_join(state_map,pet_data, by = "region")

# Graph for the percentage of people that would pay 4k to save their dog
pay4kGraph <- ggplot(state_map_pets, aes(x= long, y = lat, group = group)) +
  geom_polygon(aes(fill = spend4kPerc)) +
  scale_fill_viridis_c(option = "turbo")
pay4kGraph
```

<b>Next we'll plot the median income per state</b>

```{r}
# Joining data set with state map
state_map_med <- left_join(state_map,med_income, by = "region")

# Graph for the percentage of people that would pay 4k to save their dog
medGraph <- ggplot(state_map_med, aes(x= long, y = lat, group = group)) +
  geom_polygon(aes(fill = income)) +
  scale_fill_viridis_c(option = "turbo")
medGraph
```

<b>There is some correlation between the two maps. States with higher incomes such as California, Massachusetts, and New Jersey also have higher percentages of people that would spend \$4K to save their dogs. Similarly, states with lower income like New Mexico, Arkansas, and Oklahoma were less likely to save their dog for \$4K. This suggests that there is a correlation between median incomes and people's willingness/ability to spend \$4K to save their dog.</b>

## General Conclusion

<b>
1. Our pet ownership statistics data set had enough data to accurately answer all of our questions.

2. A better way to answer these questions may involve making multiple graphs and comparing their complexities, going more in depth.

3. Based on our findings, we would recommend decision makers to be aware of these statistics and what they mean for people and their pets. We found people’s dog devotion scores did not appear to correlate with whether or not they stayed at a job they disliked or were willing to spend $4k to save their dog. While people may have high devotion and care for their pets, it doesn’t mean they have the means to, or are willing to spend up to $4k on them or stay at a disliked job. With this, we can recommend decision makers to look more into pet ownership expenses, either before getting their pet or why the bills are costing so much in the first place. 
An important aspect to also take into consideration when making a decision involving the data would be the preference of the type of pets that exist among people. After the analysis of our first and second questions we see a strong preference for dogs over cats; which is important to consider when we look at how much devotion exists for dogs when compared to other pets in the country.

4. As stated above, there was a correlation between states incomes and their willingness to pay $4K to save their dog. This indicates that there is a corelation between the two. This indicates that even if a dog owner is very devoted to their dog, they may not be fincanically stable enough to make such a large payment, even if they wanted to.
</b>
